Data migration under a Changing Schema

### 2023-06-28 2022 Migration/articleMigration.tex
Minor

         A developer defines a new information system, $\infsys'$, which contains a data set of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.} as illustrated in figure~\ref{fig:post-migration}.

@@

         The generated script must be free of type errors, to ensure that $\sat{u}{\dataset}$ exists for every rule $u$ and data set $\dataset$.
         The proof is available \href{https://www.isa-afp.org/}{here}

@@

\subsection{Formalization}
   Let us define the instruments needed to describe the derivation of $\schema_{\migrsys}$.
\begin{definition}[union of data sets]
\label{def:graph_union_wellTyped_if_parts_wellTyped}
    \[\pair{\triples_1}{\inst_1}\cup\pair{\triples_2}{\inst_2}\ =\ \pair{\triples_1\cup\triples_2}{\inst_1\cup\inst_2}\]
\end{definition}

\begin{lemma}
\label{lemma:graph_union_wellTyped_if_parts_wellTyped}
   If $\dataset_1$ and $\dataset_2$ are data sets, then $\dataset_1\cup\dataset_2$ is a data set.
\end{lemma}

\begin{definition}[map relation names in data sets]
\label{def:map_labels_preserves_wellTypedness}
   Let $h:\Rels\rightarrow\Rels$, then
   \[\begin{array}{rcl}
      \maprel{h}{\pair{\triples}{\inst}}&=&\pair{\maprel{h}{\triples}}{\inst}\\
      \maprel{h}{\triple{a}{r}{b}}&=&\triple{a}{h(r)}{b}
   \end{array}\]
\end{definition}

\begin{lemma}
\label{lemma:map_labels_preserves_wellTypedness}
   If $\dataset$ is a data set and $h:\Rels\rightarrow\Rels$ is an injective function, then $\maprel{h}{\dataset}$ is a data set.
\end{lemma}

\begin{definition}[disjoint union of data sets]
\begin{eqnarray}
   \pair{\triples}{\inst}\sqcup\pair{\triples'}{\inst'}&=&\pair{\triples\uplus\triples'}{\inst\uplus^2\inst'}\\
      X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}
\end{eqnarray}
\end{definition}
   In a disjoint union, the instances of the two data sets are relabeled to avoid name clashes.
   In practice, we will not use 0 and 1 as labels for $x$ and $y$, but rather extend their names by prefixing.
   For example, \texttt{old.Person} and \texttt{new.Address} might be the names of concepts \texttt{Person} and \texttt{Address} from contexts \texttt{old} and \texttt{new}, respectively.
\begin{lemma}
   If $\dataset$ and $\dataset'$ are data sets, then so is $\dataset\sqcup\dataset'$.
\end{lemma}
   This lemma is being proven in Isabelle/HOL~\cite{Isabelle}. The proof is published \href{location.domain}{here}

\begin{definition}[disjoint union of schemas]
\begin{eqnarray}
   \triple{\concepts}{\rels}{\rules}\sqcup\triple{\concepts'}{\rels'}{\rules'}&=&\triple{\concepts\uplus\concepts}{\rels\uplus\rels'}{\rules\uplus\rules'}
\end{eqnarray}
\end{definition}
% \begin{definition}[disjoint union of data sets]
% \begin{eqnarray}
%    \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\dataset'\ra$}\notag\\
%    &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\instuplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
%       X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
%       X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}\\
%       X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
%       X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in Y\}\end{array}
% \end{eqnarray}
% \end{definition}
\begin{lemma}
   If $\schema$ and $\schema'$ are schemas, then so is $\schema\sqcup\schema'$.
\end{lemma}
The proof of this lemma is published \href{location.domain}{here}

To implement the disjoint union, the following relabelling functions are needed:
\begin{definition}[relabel concepts]
   \[\subst{C}{D}{A}\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ D\ \text{\bf else}\ A\]
\end{definition}
\begin{definition}[relabel concepts in triples]
   \[\subst{C}{D}{\triple{a}{r}{b}} = \triple{a}{\subst{C}{D}{r}}{b}\]
\end{definition}
\begin{definition}[relabel concepts in relations]
   \[\subst{C}{D}{(\declare{n}{A}{B})} = \declare{n}{\subst{C}{D}{A}}{\subst{C}{D}{B}}\]
\end{definition}
\begin{definition}[relabel concepts in $\inst$]
   \[a\ \subst{C}{D}{\inst}\ \subst{C}{D}{A}\ \Leftrightarrow\ a\inst A\]
\end{definition}
\begin{definition}[relabel concepts in data sets]
   \[\begin{array}{rcl}
      \rlap{$\subst{C}{D}{\pair{\triples}{\inst}}$}\\
      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\subst{C}{D}{\inst}}
   \end{array}\]
\end{definition}
   The relabeling of concepts in rules must satisfy:
\begin{equation}
   \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}
\end{equation}
\begin{definition}[relabel concepts in schemas]
   \begin{eqnarray}
      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
      &=&\begin{array}[t]{l@{}l}
         \la&\{\subst{C}{D}{c}\mid c\in\concepts\}\\
         ,&\{\subst{C}{D}{r}\mid r\in\rels\}\\
         ,&\{\subst{C}{D}{u}\mid u\in\rules\}\ \ra\notag
         \end{array}
   \end{eqnarray}
\end{definition}
\begin{definition}[relabel relations]
   \[\subst{\declare{m}{C}{D}}{m'}{{\declare{n}{A}{B}}}\ =\ \text{\bf if}\ n=m\wedge A=C\wedge B=D\ \text{\bf then}\ \declare{m'}{A}{B}\ \text{\bf else}\ \declare{m}{A}{B}\]
\end{definition}
\begin{definition}[relabel relations in triples]
   \[\subst{r'}{m'}{\triple{a}{r}{b}} = \triple{a}{\subst{r'}{m'}{r}}{b}\]
\end{definition}
\begin{definition}[relabel relations in data sets]
   \[\begin{array}{rcl}
      \rlap{$\subst{r'}{m'}{\pair{\triples}{\inst}}$}\\
      &=&\pair{\{\subst{r'}{m'}{t}\mid t\in\triples\}}{\inst}
   \end{array}\]
\end{definition}
   The relabeling of relation names in rules must satisfy:
\begin{equation}
   \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{r'}{m'}{u}}{\subst{r'}{m'}{\dataset}}
\end{equation}
\begin{definition}[relabel relations in schemas]
   \begin{eqnarray}
      \rlap{$\subst{r'}{m'}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
      &=&\triple{\concepts}{\{\subst{r'}{m'}{r} \mid r\in\rels\}}{\{\subst{r'}{m'}{u} \mid u\in\rules\}}\notag
   \end{eqnarray}
\end{definition}

   The relabeling of concepts is not trivial because of specialization.
   To illustrate this point, suppose $D\isa A$.
   Then the substitution $[C\rightarrow D]$ means that every atom $a$ that used to be of concept $C$ in the existing system, is now not only a $D$ but also an $A$.
   For example, by relabeling the concept {\tt Hotel} to {\tt Motel}, in a situation where ${\tt Motel}\isa{\tt Parking}$,
   all atoms that represent hotels might suddenly be expected to have parking lots.
   So, whereas such examples may fly in a technical sense, the developer must use renaming with care to preserve the intended semantics.

@@

   ROLE MigrationEngineer MAINTAINS migrating_takes,
                                    migrating_isPartOf,
                                    migrating_isEnrolledFor
   RULE migrating_takes : old.takes-copied_takes = -V
   RULE migrating_isPartOf : old.isPartOf-copied_isPartOf = -V
   RULE migrating_isEnrolledFor :
        old.isEnrolledFor-copied_isEnrolledFor = -V
\end{verbatim}

>>>>

         A developer defines a new information system, $\infsys'$, which contains a data set of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{In this paper, we only consider the functionality captured by rules.} as illustrated in figure~\ref{fig:post-migration}.

@@

%         The generated script must be free of type errors, to ensure that $\sat{u}{\dataset}$ exists for every rule $u$ and data set $\dataset$.
%         The proof is available \href{https://www.isa-afp.org/}{here}

@@

%SJC on Formalization: it's often better to put these inline where we first use them. Leaving this as a comment for easy copy-paste.
%\subsection{Formalization}
%   Let us define the instruments needed to describe the derivation of $\schema_{\migrsys}$.
%\begin{definition}[union of data sets]
%\label{def:graph_union_wellTyped_if_parts_wellTyped}
%    \[\pair{\triples_1}{\inst_1}\cup\pair{\triples_2}{\inst_2}\ =\ \pair{\triples_1\cup\triples_2}{\inst_1\cup\inst_2}\]
%\end{definition}
%
%\begin{lemma}
%\label{lemma:graph_union_wellTyped_if_parts_wellTyped}
%   If $\dataset_1$ and $\dataset_2$ are data sets, then $\dataset_1\cup\dataset_2$ is a data set.
%\end{lemma}
%
%\begin{definition}[map relation names in data sets]
%\label{def:map_labels_preserves_wellTypedness}
%   Let $h:\Rels\rightarrow\Rels$, then
%   \[\begin{array}{rcl}
%      \maprel{h}{\pair{\triples}{\inst}}&=&\pair{\maprel{h}{\triples}}{\inst}\\
%      \maprel{h}{\triple{a}{r}{b}}&=&\triple{a}{h(r)}{b}
%   \end{array}\]
%\end{definition}
%
%\begin{lemma}
%\label{lemma:map_labels_preserves_wellTypedness}
%   If $\dataset$ is a data set and $h:\Rels\rightarrow\Rels$ is an injective function, then $\maprel{h}{\dataset}$ is a data set.
%\end{lemma}
%
%\begin{definition}[disjoint union of data sets]
%\begin{eqnarray}
%   \pair{\triples}{\inst}\sqcup\pair{\triples'}{\inst'}&=&\pair{\triples\uplus\triples'}{\inst\uplus^2\inst'}\\
%      X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
%      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}
%\end{eqnarray}
%\end{definition}
%   In a disjoint union, the instances of the two data sets are relabeled to avoid name clashes.
%   In practice, we will not use 0 and 1 as labels for $x$ and $y$, but rather extend their names by prefixing.
%   For example, \texttt{old.Person} and \texttt{new.Address} might be the names of concepts \texttt{Person} and \texttt{Address} from contexts \texttt{old} and \texttt{new}, respectively.
%\begin{lemma}
%   If $\dataset$ and $\dataset'$ are data sets, then so is $\dataset\sqcup\dataset'$.
%\end{lemma}
%   This lemma is proven in Isabelle/HOL~\cite{Isabelle}. % TODO: point to formalization %The proof is published \href{location.domain}{here}
%
%\begin{definition}[disjoint union of schemas]
%\begin{eqnarray}
%   \triple{\concepts}{\rels}{\rules}\sqcup\triple{\concepts'}{\rels'}{\rules'}&=&\triple{\concepts\uplus\concepts}{\rels\uplus\rels'}{\rules\uplus\rules'}
%\end{eqnarray}
%\end{definition}
%% \begin{definition}[disjoint union of data sets]
%% \begin{eqnarray}
%%    \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\dataset'\ra$}\notag\\
%%    &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\instuplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
%%       X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
%%       X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}\\
%%       X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
%%       X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in Y\}\end{array}
%% \end{eqnarray}
%% \end{definition}
%\begin{lemma}
%   If $\schema$ and $\schema'$ are schemas, then so is $\schema\sqcup\schema'$.
%\end{lemma}
%The proof of this lemma is published \href{location.domain}{here}
%
%To implement the disjoint union, the following relabelling functions are needed:
%\begin{definition}[relabel concepts]
%   \[\subst{C}{D}{A}\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ D\ \text{\bf else}\ A\]
%\end{definition}
%\begin{definition}[relabel concepts in triples]
%   \[\subst{C}{D}{\triple{a}{r}{b}} = \triple{a}{\subst{C}{D}{r}}{b}\]
%\end{definition}
%\begin{definition}[relabel concepts in relations]
%   \[\subst{C}{D}{(\declare{n}{A}{B})} = \declare{n}{\subst{C}{D}{A}}{\subst{C}{D}{B}}\]
%\end{definition}
%\begin{definition}[relabel concepts in $\inst$]
%   \[a\ \subst{C}{D}{\inst}\ \subst{C}{D}{A}\ \Leftrightarrow\ a\inst A\]
%\end{definition}
%\begin{definition}[relabel concepts in data sets]
%   \[\begin{array}{rcl}
%      \rlap{$\subst{C}{D}{\pair{\triples}{\inst}}$}\\
%      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\subst{C}{D}{\inst}}
%   \end{array}\]
%\end{definition}
%   The relabeling of concepts in rules must satisfy:
%\begin{equation}
%   \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}
%\end{equation}
%\begin{definition}[relabel concepts in schemas]
%   \begin{eqnarray}
%      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
%      &=&\begin{array}[t]{l@{}l}
%         \la&\{\subst{C}{D}{c}\mid c\in\concepts\}\\
%         ,&\{\subst{C}{D}{r}\mid r\in\rels\}\\
%         ,&\{\subst{C}{D}{u}\mid u\in\rules\}\ \ra\notag
%         \end{array}
%   \end{eqnarray}
%\end{definition}
%\begin{definition}[relabel relations]
%   \[\subst{\declare{m}{C}{D}}{m'}{{\declare{n}{A}{B}}}\ =\ \text{\bf if}\ n=m\wedge A=C\wedge B=D\ \text{\bf then}\ \declare{m'}{A}{B}\ \text{\bf else}\ \declare{m}{A}{B}\]
%\end{definition}
%\begin{definition}[relabel relations in triples]
%   \[\subst{r'}{m'}{\triple{a}{r}{b}} = \triple{a}{\subst{r'}{m'}{r}}{b}\]
%\end{definition}
%\begin{definition}[relabel relations in data sets]
%   \[\begin{array}{rcl}
%      \rlap{$\subst{r'}{m'}{\pair{\triples}{\inst}}$}\\
%      &=&\pair{\{\subst{r'}{m'}{t}\mid t\in\triples\}}{\inst}
%   \end{array}\]
%\end{definition}
%   The relabeling of relation names in rules must satisfy:
%\begin{equation}
%   \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{r'}{m'}{u}}{\subst{r'}{m'}{\dataset}}
%\end{equation}
%\begin{definition}[relabel relations in schemas]
%   \begin{eqnarray}
%      \rlap{$\subst{r'}{m'}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
%      &=&\triple{\concepts}{\{\subst{r'}{m'}{r} \mid r\in\rels\}}{\{\subst{r'}{m'}{u} \mid u\in\rules\}}\notag
%   \end{eqnarray}
%\end{definition}
%   
%   The relabeling of concepts is not trivial because of specialization.
%   To illustrate this point, suppose $D\isa A$.
%   Then the substitution $[C\rightarrow D]$ means that every atom $a$ that used to be of concept $C$ in the existing system, is now not only a $D$ but also an $A$.
%   For example, by relabeling the concept {\tt Hotel} to {\tt Motel}, in a situation where ${\tt Motel}\isa{\tt Parking}$,
%   all atoms that represent hotels might suddenly be expected to have parking lots.
%   So, whereas such examples may fly in a technical sense, the developer must use renaming with care to preserve the intended semantics.

@@

   RULE ExamRule1: new.student~;new.course |- new.takes
   RULE ExamRule2: new.student~;new.course;new.isPartOf~ 
                 |- new.isEnrolledFor
\end{verbatim}
